\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-04 paper/main.tex
minor edits, wording tweaks, other fixes

A core requirement of modern scientific research is replicability.
For computational research, replicability means that code should
produce the same results, even when run on different systems.  The
standard solution to improving replicability entails packaging a
project's dependencies along with its primary code base.  Existing
solutions vary in how deeply these dependencies are specified, ranging
from virtual environments (which specify all Python package versions),
to containers (which also specify the operating system), to
virtual machines (which also specify hardware layers of the system).
Each of these existing solutions requires installing or setting up a
system for running the desired code that must be packaged alongside
the primary code base.  Here we propose an even lighter-weight
solution than virtual environments: the \texttt{davos} library.  When
used in combination with a notebook-based Python project, \texttt{davos}
library provides a mechanism for specifying (and automatically
installing) the correct package versions of the project's.  This
enables researchers to share a complete reproducible environment using
a single Jupyter notebook file.

@@

\begin{table}[H]
\begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
\hline
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Metadata value} \\
\hline
C1 & Current code version &  v0.1.1 \\
\hline
C2 & Permanent link to code/repository used for this code version & \url{https://github.com/ContextLab/davos/tree/v0.1.1} \\
\hline
C3 & Code Ocean compute capsule & \\
\hline
C4 & Legal Code License & MIT \\
\hline
C5 & Code versioning system used & git \\
\hline
C6 & Software code languages, tools, and services used & Python, JavaScript, PyPI/pip, IPython, Jupyter, Ipykernel, PyZMQ. Additional tools used for tests: pytest, Selenium, Requests, mypy, GitHub Actions \\
\hline
C7 & Compilation requirements, operating environments \& dependencies & Dependencies:~Python$>$=3.6, packaging, setuptools.~Supported OSes: MacOS, Linux, Unix-like.~Supported IPython environments: Jupyter notebooks, JupyterLab, Google Colaboratory, Binder, IDE-based notebook editors. \\
\hline
C8 & Link to developer documentation/manual & \url{https://github.com/ContextLab/davos\#readme} \\
\hline
C9 & Support email for questions & contextualdynamics@gmail.com \\
\hline
\end{tabular}
\caption{Code metadata}
\label{}
\end{table}

@@

Code sharing is a core component of the open science movement
that has inspired a full ecosystem of tools and packages.  However,
sharing code, in and of itself, does not guarantee that others will be
able to \textit{reproduce} the desired results.  For example, research
code often requires installing other software packages that extends
the language's basic functionality.  Within the Python community~\citep{vanR95},
external packages that are published in the most popular 
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despite that it is \textit{possible} to manually install the intended version numbers of
every dependency of a Python script or package, doing so may cause
conflicts within the user's computing environment that interferes with
the functionality of  \textit{other} code.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text Python scripts (.py) provide the
  most basic system for sharing raw code.  Scripts may reference
  external libraries, but those libraries are not automatically
  installed on other users' systems, nor is any version checking
  performed by default.  Jupyter notebooks (.ipynb files) comprise embedded text,
  executable code, and media (including rendered figures, code output,
  etc.).  When the \texttt{davos} library is imported into a Jupyter
  notebook, the notebook's functionality is extended to automatically
  install the required external libraries (at their correct versions,
  when specified).  Virtual environments install an isolated copy of
  Python and all required dependencies.  This typically requires
  defining a \texttt{requirements.txt} file that lists all
  dependencies (including version numbers) along with an environment
  (.yml) file that specifies how the virtual environment should be
  configured.  Containers provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to a virtual
  environment or other configurations needed to provide the necessary
  computing environment.  Containers are typically defined using
  specification files (e.g., a plain-text \texttt{dockerfile}) that
  instruct the virtualization engine regarding how to build the
  virtual environment.  Virtual machines provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary images files) must also define operating
  system-level properties of the computing environment.  Systems to
  the left of the blue vertical line entail sharing using individual
  files, with no additional installation or configuration needed to
  run the target software.  Systems to the right of the red vertical
  line provide high reproducibility by supporting precise control over
  dependencies and versioning.}
\label{fig:code-sharing}
\end{figure}

To facilitate code sharing, the Python community has developed a broad
set of approaches and tools (Fig.~\ref{fig:code-sharing}).  At one
extreme, simply publishing a set of Python scripts (.py files) may
enable others to use or gain insights into the relevant work.  Because
Python is installed by default on most modern operating systems, for
some projects this may be sufficient.  Another popular approach entails
creating JSON files, called Jupyter notebooks~\citep{KluyEtal16} that
comprise a mix of text, executable code, and embedded media.
Notebooks may call or import external scripts or libraries in order to
provide a more compact and readable experience for the users.  Each of
these systems (Python scripts and notebooks) provides a convenient
means of sharing code, with the caveat that they do not specify a
complete computing environment.  Therefore the functionality of code
shared using these systems cannot be guaranteed across different
computing environments.

At another extreme, virtual machines~\citep{Gold74, AltiEtal05, vmware} provide a complete hardware-level
simulation of the desired system.  Virtual machines are typically
fully isolated from the user's system such that installing or running
software on a virtual machine does not impact the user's primary
operating system or computing environment.  Containers~\citep[e.g.,][]{Merk14,
  KurtEtal17} provide a similar ``isolated'' experience.  Although
containerized environments do not specify hardware-level operations,
they are typically packaged with a complete operating system, in
addition to a complete copy of Python and any relevant dependencies.
Virtual environments~\citep[e.g.,][]{vanREtal14} also provide a
computing environment that is largely separated from the user's main
environment.  They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce a complete copy of the original operating system.  Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees-- at the hardware-level, operating system
level, and Python environment, respectively) that the relevant code
will run similarly for different users.  However, each of these
systems also requires installing independent tools that
can be resource intensive or burdensome to install or configure.

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the original.  Like standard Jupyter notebooks,
\texttt{davos}-enhanced notebooks allows researchers to include text,
executable code, and media within a single file.  No further setup or
installation is required, beyond what is needed to run standard
Jupyter notebooks.  And like virtual environments \texttt{davos}
provides a convenient mechanism for fully specifying (and installing,
as needed) a complete set of Python dependencies, including package versions.

@@

Like virtual environments, containers, and virtual machines, 
the \texttt{davos} library (when used in conjunction with Jupyter
notebook) provides a lightweight mechanism for sharing code and
ensuring reproducibility across users and computing environments.
Further,\texttt{davos} enables users to fully specify (and
install, as needed) any project dependencies within the same
notebook.  This provides a system whereby executable code (along with
text and media) \textit{and} code for setting up and configuring the
project dependencies, all within a single notebook file.

We designed \texttt{davos} for use in research applications.  For
example, in many cases \texttt{davos} may be used as a drop-in
replacement for more-difficult-to-set-up virtual environments,
containers, and/or virtual machines.  For researchers, this lowers barriers to sharing
code.  And by eliminating most of the setup costs of reconstructing
the researchers' computing environment, \texttt{davos} also serves to
lower barriers to entry for members of the scientific community and of
the public to \textit{benefiting} from shared code.

Beyond research applications, \texttt{davos} is also useful in
pedagogical settings.  For example, in programming courses,
instructors and students may choose to import the \texttt{davos}
library into their notebooks to provide a simple means of ensuring
their code will run on others' machines.  When combined with online
notebook-based platforms like Google Colaboratory, \texttt{davos}
provides a convenient means of managing dependencies within a
notebook, without requiring any software (beyond a web browser) to be
installed on the students' or instructors' systems.  For the same
reasons, \texttt{davos} also provides an elegant means of sharing
ready-to-run notebook-based demonstrations that install their
dependencies automatically.

In addition to the above common uses of \texttt{davos}, our work also
provides several more subtle ``advanced'' use cases.  
Whereas Python's built-in \texttt{import} statement is agnostic to
packages' version numbers, \texttt{smuggle} statements (when combined
with onion comments) are version-sensitive.  This setup enables
multiple versions of a single library to be imported within the same
notebook.  This could be useful in cases where specific features were
added or removed from a package across different versions, or in
comparing the performance or functionality of particular features
across different versions of the same package.

A second advanced use case is in providing a proof of concept of how
one can add new ``keywords'' to the Python language by leveraging the
error-handling mechanisms built into Jupyter notebooks.  This could
lead to exciting new tools that, like \texttt{davos}, extend the
Python language in useful ways.  We note that our approach to adding
the \texttt{smuggle} keyword to Python when \texttt{davos} is imported
into a notebook-based environment also carries some potential risk.
For example, a malicious user could use a similar approach to
substantially change a notebook's functionality, in
difficult-to-predict ways when a particular library was imported into
the current environment.  This highlights an important reason why
security-conscious users would be well-served to only make use of
libraries from trusted sources, or whose code is publicly available
for review.

@@

Perhaps the most exciting uses of the \texttt{davos} library are those
that we have \textit{not} yet considered or imagined.  We hope that
the Python community will find \texttt{davos} to provide a convenient
means of managing project dependencies to facilitate code sharing.  We
also hope that some of the more advanced applications of our library
might lead to new insights or discoveries.

>>>>

  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments (which specify all Python package
  versions), to containers (which also specify the operating system),
  to virtual machines (which also specify hardware layers of the
  system).  Each of these existing solutions requires installing or
  setting up a system for running the desired code that must be
  packaged alongside the primary code base.  Here we propose a
  lighter-weight solution than virtual environments: the
  \texttt{davos} library.  When used in combination with a
  notebook-based Python project, the \texttt{davos} library provides a
  mechanism for specifying (and automatically installing) the correct
  package versions of the project's dependencies.  This enables researchers to
  share a complete reproducible environment using a single Jupyter
  notebook file.

@@

\begin{table}[H]
\begin{tabular}{|l|p{6.5cm}|p{6.5cm}|}
\hline
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Metadata value} \\
\hline
C1 & Current code version &  v0.1.1 \\
\hline
C2 & Permanent link to code/repository used for this code version & \url{https://github.com/ContextLab/davos/tree/v0.1.1} \\
\hline
C3 & Code Ocean compute capsule & \\
\hline
C4 & Legal Code License & MIT \\
\hline
C5 & Code versioning system used & git \\
\hline
C6 & Software code languages, tools, and services used & Python, JavaScript, PyPI/pip, IPython, Jupyter, Ipykernel, PyZMQ. Additional tools used for tests: pytest, Selenium, Requests, mypy, GitHub Actions \\
\hline
C7 & Compilation requirements, operating environments, and
     dependencies & Dependencies:~Python $\geq 3.6$, packaging, setuptools.~Supported OSes: MacOS, Linux, Unix-like.~Supported IPython environments: Jupyter notebooks, JupyterLab, Google Colaboratory, Binder, IDE-based notebook editors. \\
\hline
C8 & Link to developer documentation/manual & \url{https://github.com/ContextLab/davos\#readme} \\
\hline
C9 & Support email for questions & contextualdynamics@gmail.com \\
\hline
\end{tabular}
\caption{Code metadata}
\label{}
\end{table}

@@

Code sharing is a core component of the open science movement that has
inspired the development of a full ecosystem of tools and packages.
However, sharing code, in and of itself, does not guarantee that
others will be able to \textit{reproduce} the desired results.  For
example, research code often requires installing other software
packages that extend the implementation language's basic
functionality.  Within the Python community~\citep{vanR95}, external
packages that are published in the most popular
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despite that it is \textit{possible} to manually install the intended
version numbers of every dependency of a Python script or package,
doing so may cause conflicts within the user's computing environment
that interfere with the functionality of \textit{other} code.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (.py files) provide the
  most basic ``system'' for sharing raw code.  Scripts may reference
  external libraries, but those libraries are not automatically
  installed on other users' systems, nor is any version checking
  performed by default.  \textbf{Jupyter notebooks} (.ipynb files) comprise embedded text,
  executable code, and media (including rendered figures, code output,
  etc.).  When the \textbf{\texttt{davos} library} is imported into a Jupyter
  notebook, the notebook's functionality is extended to automatically
  install the required external libraries (at their correct versions,
  when specified).  \textbf{Virtual environments} install an isolated copy of
  Python and all required dependencies.  This typically requires
  defining a \texttt{requirements.txt} file that lists all
  dependencies (including version numbers) along with an environment
  (.yml) file that specifies how the virtual environment should be
  configured.  \textbf{Containers} provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to a virtual
  environment or other configurations needed to provide the necessary
  computing environment.  Containers are typically defined using
  specification files (e.g., a plain-text \texttt{dockerfile}) that
  instruct the virtualization engine regarding how to build the
  virtual environment.  \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary images files) must also define operating
  system-level properties of the computing environment.  Systems to
  the left of the blue vertical line entail sharing individual
  files, with no additional installation or configuration needed to
  run the target software.  Systems to the right of the red vertical
  line provide high reproducibility by supporting precise control over
  dependencies and versioning.  Notebooks enhanced using the
  \texttt{davos} library are easily shareable and require minimal
  setup costs, while also facilitating high reproducibility by
  enabling precise control over project dependencies.}
\label{fig:code-sharing}
\end{figure}

To facilitate code sharing, the Python community has developed a broad
set of approaches and tools (Fig.~\ref{fig:code-sharing}).  At one
extreme, simply publishing a set of Python scripts (.py files) may
enable others to use or gain insights into the relevant work.  Because
Python is installed by default on most modern operating systems, for
some projects this may be sufficient.  Another popular approach
entails creating JSON files, called Jupyter
notebooks~\citep{KluyEtal16}, that comprise a mix of text, executable
code, and embedded media.  Notebooks may call or import external
scripts or libraries in order to provide a more compact and readable
experience for users.  Each of these systems (Python scripts and
notebooks) provides a convenient means of sharing code, with the
caveat that they do not specify the computing environment in which the
code is executed.  Therefore the functionality of code shared using
these systems cannot be guaranteed across different computing
environments.

At another extreme, virtual machines~\citep{Gold74, AltiEtal05,
  vmware} provide a hardware-level simulation of the desired system.
Virtual machines are typically isolated from the user's system, such
that installing or running software on a virtual machine does not
impact the user's primary operating system or computing environment.
Containers~\citep[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience.  Although containerized environments do not
specify hardware-level operations, they are typically packaged with a
complete operating system, in addition to a complete copy of Python
and any relevant package dependencies.  Virtual
environments~\citep[e.g.,][]{vanREtal14} also provide a computing
environment that is largely separated from the user's main
environment.  They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce an operating system for the runtime environment.  Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees-- at the hardware level, operating
system level, and Python environment level, respectively) that the
relevant code will run similarly for different users.  However, each
of these systems also relies on additional software that can be
resource intensive or burdensome to install or configure.

We designed \texttt{davos} to occupy a ``sweet spot'' between these
extremes.  \texttt{davos} is a notebook-installable package that adds
functionality to the default notebook experience.  Like standard
Jupyter notebooks, \texttt{davos}-enhanced notebooks allows
researchers to include text, executable code, and media within a
single file.  No further setup or installation is required, beyond
what is needed to run standard Jupyter notebooks.  And like virtual
environments \texttt{davos} provides a convenient mechanism for fully
specifying (and installing, as needed) a complete set of Python
dependencies, including package versions.

@@

Like virtual environments, containers, and virtual machines, the
\texttt{davos} library (when used in conjunction with Jupyter
notebooks) provides a lightweight mechanism for sharing code and
ensuring reproducibility across users and computing environments
(Fig.~\ref{fig:code-sharing}).  Further, \texttt{davos} enables users
to fully specify (and install, as needed) any project dependencies
within the same notebook.  This provides a system whereby executable
code (along with text and media) \textit{and} code for setting up and
configuring the project dependencies, may be combined within a single
notebook file.

We designed \texttt{davos} for use in research applications.  For
example, in many settings \texttt{davos} may be used as a drop-in
replacement for more-difficult-to-set-up virtual environments,
containers, and/or virtual machines.  For researchers, this lowers
barriers to sharing code.  By eliminating most of the setup costs
of reconstructing the original researchers' computing environment,
\texttt{davos} also lowers barriers to entry for members of
the scientific community and the public who seek to \textit{benefit}
from shared code.

Beyond research applications, \texttt{davos} is also useful in
pedagogical settings.  For example, in programming courses,
instructors and students may import the \texttt{davos} library into
their notebooks to provide a simple means of ensuring their code will
run on others' machines.  When combined with online notebook-based
platforms like Google Colaboratory, \texttt{davos} provides a
convenient way to manage dependencies within a notebook, without
requiring any software (beyond a web browser) to be installed on the
students' or instructors' systems.  For the same reasons,
\texttt{davos} also provides an elegant means of sharing ready-to-run
notebook-based demonstrations that install their dependencies
automatically.

Our work also has several more subtle ``advanced'' use cases and
potential impacts.  Whereas Python's built-in \texttt{import}
statement is agnostic to packages' version numbers, \texttt{smuggle}
statements (when combined with onion comments) are version-sensitive.
This enables multiple versions of a single library to be
imported within the same notebook.  This could be useful in cases
where specific features were added or removed from a package across
different versions, or in comparing the performance or functionality
of particular features across different versions of the same package.

A second advanced use case is in providing a proof-of-concept of how
one can add new ``keywords'' to the Python language by leveraging the
error-handling mechanisms.  This could lead to exciting new tools
that, like \texttt{davos}, extend the Python language in useful ways.
We note that our approach to adding the \texttt{smuggle} keyword to
Python when \texttt{davos} is imported into a notebook-based
environment also has the potential to be exploited for more nefarious
purposes.  For example, a malicious user could use a similar approach
(e.g., in a different library) to substantially change a notebook's
functionality by adding new \textit{unexpected} keyword-like objects
(e.g., based around common typos).  This could lead to
difficult-to-predict changes in a notebook's behavior once the
malicious library was imported.  This highlights an important reason
why security-conscious users would be well-served to only make use of
libraries from trusted sources, or whose code is publicly available
for review.

@@

The \texttt{davos} library supports reproducible research by providing
a novel lightweight system for sharing notebook-based code.  But
perhaps the most exciting uses of the \texttt{davos} library are those
that we have \textit{not} yet considered or imagined.  We hope that
the Python community will find \texttt{davos} to provide a convenient
means of managing project dependencies to facilitate code sharing.  We
also hope that some of the more advanced applications of our library
might lead to new insights or discoveries.
